Determinism, Complexity, and Predictability in Computer Performance

### 2014-07-30 model.tex
more cleanup

  \hline\hline
  \end{tabular*}
  \end{center}
 \label{tab:error}
  \end{table*}%
In view of the discussion at the end of the previous paragraph, the
fact that the values in the second-from-\alert{left} column are not
identically equal to 1.00 may be somewhat surprising.  This can happen
due to differences between in-sample and out-of-sample forecasting;
the last 10\% of \svdthree, for instance---the out-of-sample
signal---was easier, on the average, for RW to predict than the first
90\%. \alert{[[Liz to do: add pointer to Josh's paragraph for deeper
      analysis]]}.

@@

Normalization is, as is often the case, a double-edged sword.  Here,
it facilitates comparisons across signals of different types and
lengths, but its particular pathologies must be taken into account, as
discussed at more length in Section~\ref{sec:results}.

>>>>

  \hline\hline
  \end{tabular*}
  \end{center}
 \label{tab:error}
  \end{table*}%
In view of the discussion at the end of the previous paragraph, the
fact that the values in the second-from-\alert{left} column are not
identically equal to 1.00 may be somewhat surprising.  This can happen
due to differences between in-sample and out-of-sample forecasting;
the last 10\% of \svdthree, for instance---the out-of-sample
signal---was more amenable to random-walk prediction, on the average,
than the first 90\%.  See Section~\ref{sec:results} for a deeper
discussion of this effect.

@@

Normalization is, as is often the case, a double-edged sword.  Here,
it facilitates comparisons across signals of different types and
lengths, but its particular pathologies must be taken into account
when analyzing the results, as discussed at more length in
Section~\ref{sec:results}.

### 2014-07-30 results.tex
more cleanup

Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
corresponding MASE values of the \emph{best} prediction for each of
the 120 time series in this study.
\begin{figure}
  \centering
  \includegraphics[width=1.1\columnwidth]{figs/new_prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each time series.  The solid
    curve is a least-squares log fit of these points.
% 
% of all the points except those from {\tt dgesdd$_1$}, which we have
% excluded for reasons explained in the text.
% 
The dashed curves reflect the standard deviation of the model in its
parameter space.  The meaning of the shaded region is explained in the
text.}
  \label{fig:wpe_vs_mase_best}
\end{figure}
% 
% As the plot shows, the best-case prediction error for each trace from
% these eight systems is roughly proportional to the log of the weighted
% permutation entropy, which is consistent with our first conjecture.
% 
There is an obvious upward trend, which is consistent with the notion
that there is a pattern in the WPE-MASE relationship.  However, a
simple linear fit is a bad idea here.  First, any signal with zero
entropy should be perfectly predictable (i.e., MASE $\approx 0$), so
any curve fitted to these data should pass through the origin.
Moreover, WPE does not grow without bound, so one would expect the
patterns in the WPE-MASE pairs to reach some sort of asymptote.  For
these reasons, we chose to fit a function of the form $y = a \log(b x
+ 1)$ to these points, with $y =$ WPE and $x=$ MASE.  \alert{can we
  cite a figure of merit for this fit?  we also need to say what $a$
  and $b$ were.}  The solid curve in the figure shows this fit; the
dashed curves show the standard deviation of this model in its
parameter space: i.e., $y = a \log(b x + 1)$ with $\pm$ one standard
deviation on each of the two parameters.  Points that fall within this
deviation volume \alert{(light grey)} correspond to predictions that
are comparable to the best ones found in this study; points that fall
\emph{above} that volume \alert{(dark grey)} are better still.  We
chose to truncate the shaded region because of a subtle point
regarding the MASE of an ideal predictor, which should not be larger
than 1 unless the training and test signals are different.  This is
discussed at more length below.
%[[make sure that's true]]

@@

The main thing to note here, however, is not the details of the shapes
of the clusters, but rather their positions in the four plots:
specifically, the fact that many of them are to the right of and/or
below the dashed curve that identifies the boundary of the shaded
region.  \emph{These predictions are not as good as our heuristic
  suggests they could be.}  Focusing in on any single cluster of
points makes this clear: LMA works best for \svdsix, for instance,
followed by the random-walk prediction method, then \arima and \naive.
When a practioner calculates an WPE vs. MASE value that is outside the
shaded region, that should suggest to her that the prediction method
is not well matched to the task at hand: that is, the time series has
more predictive structure than the method is able to use.
% 
% The idea behind the shaded region is to help practitioners identify
% when there is predictive structure in the data that a given method is
% not exploiting (viz., \arima on \svdsix)---i.e., that one could do
% better with another forecast strategy.
% 
In this case---e.g., \arima on \svdsix---she should try a different
method.  The position of the LMA cluster for \svdsix, on the other
hand, reflects this method's ability to capture and exploit the
structure that is present in this signal and suggests to the
practitioner that [[she really can't do any better]].  The following
discussion addresses the details that underlie these claims.
% 
% The dashed curve in Figures~\ref{fig:wpe_vs_mase_best}
% and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
% argument at the end of the previous paragraph.  
%

@@

The WPE of \gcc is higher still: $0.943 \pm 0.001$.  This system
transmits very little information forward in time and provides almost
no structure for prediction methods to work with.  Here, the
random-walk predictor is the winner, with a MASE of $1.1381 \pm
0.011$.  This makes sense; in a fully complex signal, the random-walk
will always be best, assuming that the training signal accurately
reflects the overall behavior of the out-of-sample time series.
\alert{feels like we should justify this a bit}
% 
% The \naive ~method works fairly well too: $1.797 \pm 0.010$, simply
% because its calculation of the mean filters out the noise in the
% signal.
% 
Since fitting a hyperplane using least squares should filter out some
of the noise in the signal, the fact that LMA outperforms \arima
($1.530 \pm 0.021$ vs. $1.837 \pm 0.016$) may be somewhat
counterintuitive.  However, the small amount of predictive structure
that is present in this signal is nonlinear (cf.,~\cite{mytkowicz09}),
and LMA is designed to capture and exploit that kind of structure.
\alert{this is still true, no?  or should we just say ``see, LMA does
  better as one would expect because the signal is nonlinear?''}

@@

Again, the curves in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_best} were determined from a finite set of
methods and data.  [[But we worked hard to make these experiments
    representative and comprehensive: forecast methods from simple to
    sophisticated; time-series data from a system whose behavior spans
    the dynamical behavior space.]]  While we are cautiously optmistic
about its generality, more exploration will be required to firm up
that conclusion.  Our preliminary work along those lines shows that
data from the Henon map \cite{henon}, the Lorenz system \cite{lorenz},
and the SFI A data set \cite{sfi-data} all fall within the one-$\sigma$
volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} region, as do various nonlinear
transformations of \svdtwo and \svdsix.

Of course, the geometry of the curves and bounds do not hold for other
error metrics.  The positions of the WPE vs. error points---and the
reasons to choose a particular function to fit to them, or a
theoretical bound on the meaning of the results---may change if one
calculates error in a different way.  The results presented here only
hold for the MASE error metric.  Even though this metric has its
weaknesses, we concur with \cite{MASE} that it is an effective way to
compare prediction error across time-series data of different lengths
and different scales.  [[Point to that paper for extensive evaluation
    and comparison wrt other metrics.]]

>>>>

Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
corresponding MASE values of the \emph{best} prediction for each of
the 120 time series in this study.
\begin{figure}
  \centering
  \includegraphics[width=1.1\columnwidth]{figs/new_prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each time series.  The solid
    curve is a least-squares log fit of these points.
% 
% of all the points except those from {\tt dgesdd$_1$}, which we have
% excluded for reasons explained in the text.
% 
The dashed curves reflect the standard deviation of the model in its
parameter space.  The meaning of the shaded region is explained in the
text.}
  \label{fig:wpe_vs_mase_best}
\end{figure}
% 
% As the plot shows, the best-case prediction error for each trace from
% these eight systems is roughly proportional to the log of the weighted
% permutation entropy, which is consistent with our first conjecture.
% 
There is an obvious upward trend, which is consistent with the notion
that there is a pattern in the WPE-MASE relationship.  However, a
simple linear fit is a bad idea here.  First, any signal with zero
entropy should be perfectly predictable (i.e., MASE $\approx 0$), so
any curve fitted to these data should pass through the origin.
Moreover, WPE does not grow without bound, so one would expect the
patterns in the WPE-MASE pairs to reach some sort of asymptote.  For
these reasons, we chose to fit a function of the form $y = a \log(b x
+ 1)$ to these points, with $y =$ WPE and $x=$ MASE.  \alert{can we
  cite a figure of merit for this fit?  we also need to say what $a$
  and $b$ were.}  The solid curve in the figure shows this fit; the
dashed curves show the standard deviation of this model in its
parameter space: i.e., $y = a \log(b x + 1)$ with $\pm$ one standard
deviation on each of the two parameters.  Points that fall within this
deviation volume \alert{(light grey)} correspond to predictions that
are comparable to the best ones found in this study; points that fall
\emph{above} that volume \alert{(dark grey)} are better still.  We
chose to truncate the shaded region because of a subtle point
regarding the MASE of an ideal predictor, which should not be larger
than 1 unless the training and test signals are different.  This is
discussed at more length below.

@@

The main thing to note here, however, is not the details of the shapes
of the clusters, but rather their positions in the four plots:
specifically, the fact that many of them are to the right of and/or
below the dashed curve that identifies the boundary of the shaded
region.  \emph{These predictions are not as good as our heuristic
  suggests they could be.}  Focusing in on any single cluster of
points makes this clear: LMA works best for \svdsix, for instance,
followed by the random-walk prediction method, then \arima and \naive.
When a practioner calculates an WPE vs. MASE value that is outside the
shaded region, that should suggest to her that the prediction method
is not well matched to the task at hand: that is, the time series has
more predictive structure than the method is able to use.
% 
% The idea behind the shaded region is to help practitioners identify
% when there is predictive structure in the data that a given method is
% not exploiting (viz., \arima on \svdsix)---i.e., that one could do
% better with another forecast strategy.
% 
In this case---e.g., \arima on \svdsix---she should try a different
method.  The position of the LMA cluster for \svdsix, on the other
hand, reflects this method's ability to capture and exploit the
structure that is present in this signal.  WPE vs. MASE values like
this, which fall in the shaded region, should suggest to the
practitioner that the prediction method is well-suited to the task.
The following discussion lays out the details that underlie these
claims.
% 
% The dashed curve in Figures~\ref{fig:wpe_vs_mase_best}
% and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
% argument at the end of the previous paragraph.  
%

@@

The WPE of \gcc is higher still: $0.943 \pm 0.001$.  This system
transmits very little information forward in time and provides almost
no structure for prediction methods to work with.  Here, the
random-walk predictor is the winner, with a MASE of $1.1381 \pm
0.011$.  This makes sense; in a fully complex signal, the random-walk
will always be best, assuming that the training signal accurately
reflects the overall behavior of the out-of-sample time series.
\alert{feels like we should justify this a bit}
% 
% The \naive ~method works fairly well too: $1.797 \pm 0.010$, simply
% because its calculation of the mean filters out the noise in the
% signal.
% 
Since fitting a hyperplane using least squares should filter out some
of the noise in the signal, the fact that LMA outperforms \arima
($1.530 \pm 0.021$ vs. $1.837 \pm 0.016$) may be somewhat
counterintuitive.  However, the small amount of predictive structure
that is present in this signal is nonlinear (cf.,~\cite{mytkowicz09}),
and LMA is designed to capture and exploit that kind of structure.
\alert{this is still true, no?  or should we just say ``see, LMA does
  better as one would expect because the signal is nonlinear?''  also,
  we need to talk about the fact that the best of these points, and of
  \svdtwo, is past the MASE wall and not in the shaded region.}

@@

Again, the curves in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_best} were determined from a finite set of
methods and data.  We put a lot of thought and effort into making
these experiments upon which these results are based both
representative and comprehensive.  The forecast methods involved range
from the simple to the sophisticated; the time-series data was sampled
from a system whose behavior spans the dynamical behavior space.
While we are cautiously optimistic about the generality of our
conclusions, more exploration will be required before we can make any
such claim.  Our preliminary work along those lines shows that data
from the Henon map \cite{henon}, the Lorenz system \cite{lorenz}, and
the SFI A data set \cite{sfi-data} all fall within the one-$\sigma$
volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} region, as do various nonlinear
transformations of \svdtwo and \svdsix.

Of course, the geometry of these curves and bounds do not necessarily
extend to other error metrics.  The positions of the WPE vs. error
points---and the reasons to choose a particular function to fit to
them or impute theoretical bounds on the meaning of the results---may
change if one calculates error using a procedure other than the one
described in Section~\ref{sec:accuracy}.  The MASE error metric, as
discussed above, has its weaknesses.  Even so, we concur with
\cite{MASE} that it is an effective way to compare prediction error
across time-series data of different lengths and different scales, and
we are convinced by the extensive evaluations that are offered in that
paper, as well as the comparisons of MASE to other metrics, that its
strengths far outweigh its weaknesses.
